A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-26 sections/Metatheory.tex
4.1 revision, wip

This section presents the metatheory of the algorithmic system
presented in the previous section. We show that three main results hold: 
\emph{soundness}, \emph{completeness} and \emph{decidability}.
These three results have been mechanically formalized and proved in the 
Abella theorem prover~\cite{}.

\subsection{Transfer to the Declarative System}
%\jimmy{The transfer rule outputs declarative judgment chains, which are later processed by another declarative relation, handling all the guessing.}

\begin{figure}[t]
\framebox{$\Gm \sto \Om$} $\Gm$ transfers to $\Om$.
\begin{gather*}
\inferrule*[right=$\mathtt{{\sto}}\Om$]
{~}
{\Om \sto \Om}
\quad
\inferrule*[right=$\mathtt{{\sto}\al}$]
{\Om\vdash\tau \\ \Om,[\tau/\al]\Gm \sto \Om}
{\Om,\al,\Gm \sto \Om}
\end{gather*}
\caption{Transfer rules}
\label{fig:trans}
\end{figure}

To aid formalizing the correspondence between the declarative and
algorithmic systems, we use a relation that transfers algorithmic
worklist judgements into corresponding declarative worklist judgements. In essence this
judgment instantiates every existential variable with a well-formed monotype.

The transfer rules are shown in Figure~\ref{fig:trans}.
Rule $\mathtt{{\sto}\al}$ replaces the first existential variable with a well-scoped monotype and
repeats the process in the resulting worklist.
Note that $\Om$, as a declarative worklist, contains no existential
variables. Therefore by recursively applying this rule, the
algorithmic worklist will finally become a declarative one.
In order to maintain well-formedness,
the substitution should affect all the judgments and term variable bindings in the scope of $\al$.

The transfer $\Gm\sto\Om$ is not deterministic.
% \bruno{Do you mean here
%   ``is not deterministic''? Would that be more appropriate?}
On the one hand, there are infinite possibilities to instantiate an existential variable,
thus a judment $\Gm$ could be transferred to different $\Om$'s.
On the other hand, any valid monotype in $\Om$ can be represented by an existential variable in $\Gm$,
thus different $\Gm$'s could be transferred to a single $\Om$.

@@

\begin{figure}[t]
\framebox{$\|\Om\|$} Judgment erasure for $\Om$, returns a declarative context $\Psi$.
\begin{gather*}
\begin{aligned}
\|\nil\| &= \nil\\
\|\Om,a\| &= \|\Om\|, a\\
\|\Om,x:A\| &= \|\Om\|, x:A\\
\|\Om\Vdash\jg\| &= \|\Om\|
\end{aligned}
\end{gather*}

\framebox{$\Om \rto \Om'$} $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \|\Om\| \vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \|\Om\| \vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \|\Om\| \vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \|\Om\| \vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing.\bruno{I think your names for the
  two relations/definitions aren't great. I think what you're defining
here should be called transfer. The previous relation should probably
be called something different. Maybe something like Declarative
worklist instantiation/realization for the previous relation?}}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} relates the worklist-style declarative system
to the original declarative system.
A successful reduction means a valid variable declaration,
a valid subtyping or checking judgment, or a successful guessing for inferences.
Additionally, judgment chains are unfolded to match the original declarative rules.
Although $\Om$ is defined differently from $\Psi$,
our declarative worklist $\Om$ contains a ordered collection of variable definitions,
which naturally fits declarative subtyping and typing.
\bruno{I think that to make things precise what you want to show is
  that you can always define a erasure function that drops the
  judgements from the worklist, and retains only the variable
  declarations (thus becomes a declarative environment). Isn't this
  what you do in your formalization?}
\jimmy{TODO: a erasure function would make things more clear...}

>>>>

This section presents the metatheory of the algorithmic system
presented in the previous section. We show that three main results hold: 
\emph{soundness}, \emph{completeness} and \emph{decidability}.
These three results have been mechanically formalized and proved in the 
Abella theorem prover~\cite{AbellaDesc}.

%-------------------------------------------------------------------------------
\subsection{Declarative Worklist}
%\jimmy{The transfer rule outputs declarative judgment chains, which are later processed by another declarative relation, handling all the guessing.}

\begin{figure}[t]
\begin{gather*}
\begin{aligned}
\text{Declarative worklist}\qquad&\Om &::=&\quad \nil \mid \Om, a \mid \Om, x: A \mid \Om \Vdash \jg
\end{aligned}
\end{gather*}
\hfill \framebox{$\Gm \sto \Om$} \hfill $\Gm$ instantiates to $\Om$.
\begin{gather*}
\inferrule*[right=$\mathtt{{\sto}}\Om$]
{~}
{\Om \sto \Om}
\quad
\inferrule*[right=$\mathtt{{\sto}\al}$]
{\Om\vdash\tau \\ \Om,[\tau/\al]\Gm \sto \Om}
{\Om,\al,\Gm \sto \Om}
\end{gather*}
\caption{Declarative Worklists and Instantiation}
\label{fig:trans}
\end{figure}

To aid formalizing the correspondence between the declarative and algorithmic
systems, we introduce the notion of declarative worklist $\Om$, define in
Figure~\ref{fig:trans}. A declarative worklist $\Om$ has the same structure as an
algorithmic worklist $\Gm$, but does not contain any extistential variables $\al$.

The relation $\Gm \sto \Om$ expresses that the algorithmic worklist $\Gm$ can 
be instiantiated to the declarative worklist $\Om$, namely by appropriately instantiating
all exitential variables $\al$ in $\Gm$ with well-scoped monotypes $\tau$.
The rules of this instantiation relation are shown in Figure~\ref{fig:trans} too.
Rule $\mathtt{{\sto}\al}$ replaces the first existential variable with a well-scoped monotype and
repeats the process on the resulting worklist until no existential variable remains
and thus the algorithmic worklist has become a declarative one.
In order to maintain well-scopedness
the substitution is applied to all the judgments and term variable bindings in the scope of $\al$.

Observe that the instantiation $\Gm\sto\Om$ is not deterministic.
% \bruno{Do you mean here
%   ``is not deterministic''? Would that be more appropriate?}
From left to right, there are infinitely many possibilities to instantiate an existential variable and
thus infinitely many declarative worklists that one can get from an algorithmic one.
In the other direction, any valid monotype in $\Om$ can be abstracted to an
existential variable in $\Gm$. Thus different $\Gm$'s can be instantiated to the same
$\Om$.

@@

\begin{figure}[t]
\hfill \framebox{$\|\Om\|$} \hfill Judgment erasure.
\begin{gather*}
\begin{aligned}
\|\nil\| &= \nil\\
\|\Om,a\| &= \|\Om\|, a\\
\|\Om,x:A\| &= \|\Om\|, x:A\\
\|\Om\Vdash\jg\| &= \|\Om\|
\end{aligned}
\end{gather*}

\hfill \framebox{$\Om \rto \Om'$} \hfill $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \|\Om\| \vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \|\Om\| \vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \|\Om\| \vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \|\Om\| \vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing.\bruno{I think your names for the
  two relations/definitions aren't great. I think what you're defining
here should be called transfer. The previous relation should probably
be called something different. Maybe something like Declarative
worklist instantiation/realization for the previous relation?}}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} defines a declarative version $\Om \sto \Om'$ of
the algorithmic reduction rules, which relates the declarative worklists to the
declarative type system, by checking every judgement entry in the worklist with
the help of the corresponding conventional declarative judgement. Recall that
the worklist consists of an interleaving of the typing context with judgments.
By means of the auxiliary function $\|\Om\|$ we can extracts the typing context
out of it, for use in the conventional declarative judgments. What is does to
obtain the typing context is simply erase the judgment entries from the
worklist.
